---
title: "ASE embedding Geometry"
author: "Tianyi Chen"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r, message=FALSE,warning=FALSE}
library(Matrix)
library(igraph)
library(mclust)
library(randnet)
library(irlba)
library(ggplot2)
library(MASS)
library(mvtnorm)
library(dplyr)
library(RColorBrewer)
library(readr)
library(patchwork)  
edge_list = read_csv("~/Desktop/Research /PhDresearch/Hopkins_Organoid/MO VS SO_2025_May_graphs/adjacency_edges_May_31_2025_ecr_results_no_window.csv")
filenames = unique(edge_list$File)
filenames = filenames[-2]
```

```{r}
extract_path_part <- function(filepath) {
  # Use sub() to find the pattern and return only that part
  sub(".*/(M\\d+/[A-Za-z]+?/\\d+)/.*", "\\1", filepath)
}
full.ase <- function(A, d, diagaug=TRUE, doptr=FALSE) {
  require(irlba)
  
  # doptr
  if (doptr) {
    g <- ptr(A)
    A <- g[]
  } else {
    A <- A[]
  }
  A = as.matrix(A)
  
  # diagaug
  if (diagaug) {
    diag(A) <- rowSums(A) / (nrow(A)-1)
  }
  
  A.svd <- irlba(A,d)
  Xhat <- A.svd$u %*% diag(sqrt(A.svd$d))
  Xhat.R <- NULL
  
  if (!isSymmetric(A)) {
    Xhat.R <- A.svd$v %*% diag(sqrt(A.svd$d))
  }
  
  return(list(eval=A.svd$d, Xhat=Matrix(Xhat), Xhat.R=Xhat.R))
}



full.lse <- function(A, d) {
  deg.seq <- rowSums(A)
  L.svd <- irlba(A/sqrt(outer(deg.seq,deg.seq)),d)
  Xhat <- L.svd$u %*% diag(sqrt(L.svd$d))
  Xhat.R <- NULL
  
  if (!isSymmetric(A)) {
    Xhat.R <- L.svd$v %*% diag(sqrt(L.svd$d))
  }
  
  return(list(eval=L.svd$d, Xhat=(Xhat), Xhat.R=(Xhat.R)))
}

getElbows <- function(dat, n = 3, threshold = FALSE, plot = TRUE, main="") {
  if (is.matrix(dat)) {
    d <- sort(apply(dat,2,sd), decreasing=TRUE)
  } else {
    d <- sort(dat,decreasing=TRUE)
  }
  
  if (!is.logical(threshold))
    d <- d[d > threshold]
  
  p <- length(d)
  if (p == 0)
    stop(paste("d must have elements that are larger than the threshold ",
               threshold), "!", sep="")
  
  lq <- rep(0.0, p)                     # log likelihood, function of q
  for (q in 1:p) {
    mu1 <- mean(d[1:q])
    mu2 <- mean(d[-(1:q)])              # = NaN when q = p
    sigma2 <- (sum((d[1:q] - mu1)^2) + sum((d[-(1:q)] - mu2)^2)) /
      (p - 1 - (q < p))
    lq[q] <- sum( dnorm(  d[1:q ], mu1, sqrt(sigma2), log=TRUE) ) +
      sum( dnorm(d[-(1:q)], mu2, sqrt(sigma2), log=TRUE) )
  }
  
  q <- which.max(lq)
  if (n > 1 && q < (p-1)) {
    q <- c(q, q + getElbows(d[(q+1):p], n-1, plot=FALSE))
  }
  
  if (plot==TRUE) {
    if (is.matrix(dat)) {
      sdv <- d # apply(dat,2,sd)
      plot(sdv,type="b",xlab="dim",ylab="stdev",main=main)
      points(q,sdv[q],col=2,pch=19)
    } else {
      plot(dat, type="b",main=main)
      points(q,dat[q],col=2,pch=19)
    }
  }
  
  return(q)
}

```


In this document i will present the ASE embedding results for both datasets.For all graphs i will symmetrize the graph and then find the largest connected component and then apply ASE. 

## Summary of the data.

Single organoid plate – M08438 

Multi-organoid plate – M07359 

M07357--???

There are mysterious chip M07357/000388 only well004 contains 52 edges, all are removed after filter. Thus M07357/000388 is not stored in the edge list csv file. 

Now 15 files left.
Among them 3 are completely empty either no edges at all or after filter. 
For chip M08438
000396 doesn't have edges for all wells. Thus not stored in the edge list csv file. 
For chip M07359
000384 only have 2 edges in well 002 but after filter adj matrix has no edge thus not stored in the edge list file. 
000391 doesn't have edges for all wells. Thus not stored in the edge list csv file. 

Now remain 12 files in the edgelist file. Among them 2 are almost empty. 
000393 only well004 has 34 edges;
000386 only well000 and well002 have 2 edges in both. 


**However, the recording in M07359 does show clear geometry property.** Others are like the noisy version of it. 

## Embedding geometry for the new MO vs SO dataset . 
```{r,message=FALSE}
library(igraph)
library(Matrix)
library(iGraphMatch) # Assuming full.ase and getElbows are from this package
library(scatterplot3d)

for (file in filenames) {
    par(mfrow = c(2, 3), mar = c(3, 3, 3, 1), oma = c(4, 0, 3, 0))
    
    # Define the list of wells to iterate over
    well_list <- paste0("well", sprintf("%03d", 0:5))
    
    # --- Main Loop ---
    for (well in well_list) {
      
      # 1. Subset the data for the current file and well
      subset_data <- subset(edge_list, File == file & Well == well)
      
      # Initial check for any data at all
      if (nrow(subset_data) == 0) {
        plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
        title(main = well)
        text(1, 1, "No data available", col = "red")
        next # Move to the next well
      }
      
      # --- Graph Construction and ASE ---
      # This block is wrapped in tryCatch to handle potential errors
      tryCatch({
        # Create the adjacency matrix
        n_rows <- n_cols <- unique(subset_data$dim)
        adj <- sparseMatrix(
          i    = subset_data$Row + 1,
          j    = subset_data$Column + 1,
          dims = c(n_rows, n_cols)
        )
        
        # Create the graph and find the largest connected component (LCC)
        g <- graph_from_adjacency_matrix(adj, mode = "undirected", diag = FALSE)
        comps <- components(g, mode = "weak")
        big <- which.max(comps$csize)
        g_lcc <- induced_subgraph(g, V(g)[comps$membership == big])
        
        # 2. Check if the size of the LCC is less than 11
        if (vcount(g_lcc) < 11) {
          # If so, plot the specified message and skip the analysis
          plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
          title(main = well)
          text(1, 1, "Size of largest\nconnected component < 11", col = "red", cex = 0.9)
        } else {
          # --- Proceed with ASE only if LCC size is 11 or greater ---
          
          # Perform Adjacency Spectral Embedding (ASE) on the LCC
          ase <- full.ase(g_lcc, 10)
          
          # Find the elbow to determine the embedding dimension
          elb <- getElbows(ase$eval, plot = F)
          
          if (length(elb) < 2) {
            stop("Embedding dimension is less than 2.")
          }
          
          embedding_dim <- elb[2]
          
          # --- Conditional Plotting ---
          
          # 3. If the embedding dimension is 2, create a 2D plot
          if (embedding_dim == 2) {
            plot(ase$Xhat[, 1], ase$Xhat[, 2], 
                 main = well, 
                 xlab = "Dimension 1", 
                 ylab = "Dimension 2",
                 pch = 16,
                 col = "blue")
          } 
          # 4. If the embedding dimension is 3 or more, create a 3D plot
          else if (embedding_dim >= 3) {
            scatterplot3d(
              x = ase$Xhat[, 1], 
              y = ase$Xhat[, 2], 
              z = ase$Xhat[, 3],
              main = well,
              xlab = "Dim 1",
              ylab = "Dim 2",
              zlab = "Dim 3",
              color = "blue",
              pch = 16,
              type = "p",
              grid = TRUE,
              box = TRUE
            )
          } 
          # Handle cases where the dimension is less than 2
          else {
            plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
            title(main = well)
            text(1, 1, paste("Embedding dim < 2\n(dim =", embedding_dim, ")"), col = "orange")
          }
        }
        
      }, error = function(e) {
        # If any other error occurs, plot a placeholder with the error message
        plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
        title(main = well)
        text(1, 1, "An error occurred", col = "red", cex = 0.8)
        message(paste("Error processing", well, ":", e$message))
      })
    }
    
    # --- Add the Main Title to the Entire Figure ---
    # Use mtext() to write the extracted path part in the outer margin (top)
    mtext(extract_path_part(file), 
          side = 3,       # 3 = top
          line = 1,       # Position in the margin
          outer = TRUE,   # Use the outer margin
          cex = 1.5,      # Character expansion (font size)
          font = 2)       # Font style (2 = bold)
    
    # Optional: Reset the plotting device layout to default
    # par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
    
        
}

```

## For comparison, this is embedding geometry for first 3vs 1 dataset.


```{r,message=FALSE,warning=FALSE}
edge_list = read.csv('/Users/tianyichen/Desktop/Research /PhDresearch/Hopkins_Organoid/Codes/R/adjacency_edges.csv')
filenames = unique(edge_list$File)


for (file in filenames) {
  par(mfrow = c(2, 3), mar = c(3, 3, 3, 1), oma = c(4, 0, 3, 0))
  
  # Define the list of wells to iterate over
  well_list <- paste0("well", sprintf("%03d", 0:5))
  
  # --- Main Loop ---
  for (well in well_list) {
    
    # 1. Subset the data for the current file and well
    subset_data <- subset(edge_list, File == file & Well == well)
    
    # Initial check for any data at all
    if (nrow(subset_data) == 0) {
      plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
      title(main = well)
      text(1, 1, "No data available", col = "red")
      next # Move to the next well
    }
    
    # --- Graph Construction and ASE ---
    # This block is wrapped in tryCatch to handle potential errors
    tryCatch({
      # Create the adjacency matrix
      n_rows <- n_cols <- unique(subset_data$dim)
      adj <- sparseMatrix(
        i    = subset_data$Row + 1,
        j    = subset_data$Column + 1,
        dims = c(n_rows, n_cols)
      )
      
      # Create the graph and find the largest connected component (LCC)
      g <- graph_from_adjacency_matrix(adj, mode = "undirected", diag = FALSE)
      comps <- components(g, mode = "weak")
      big <- which.max(comps$csize)
      g_lcc <- induced_subgraph(g, V(g)[comps$membership == big])
      
      # 2. Check if the size of the LCC is less than 11
      if (vcount(g_lcc) < 11) {
        # If so, plot the specified message and skip the analysis
        plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
        title(main = well)
        text(1, 1, "Size of largest\nconnected component < 11", col = "red", cex = 0.9)
      } else {
        # --- Proceed with ASE only if LCC size is 11 or greater ---
        
        # Perform Adjacency Spectral Embedding (ASE) on the LCC
        ase <- full.ase(g_lcc, 10)
        
        # Find the elbow to determine the embedding dimension
        elb <- getElbows(ase$eval, plot = F)
        
        if (length(elb) < 2) {
          stop("Embedding dimension is less than 2.")
        }
        
        embedding_dim <- elb[2]
        
        # --- Conditional Plotting ---
        
        # 3. If the embedding dimension is 2, create a 2D plot
        if (embedding_dim == 2) {
          plot(ase$Xhat[, 1], ase$Xhat[, 2], 
               main = well, 
               xlab = "Dimension 1", 
               ylab = "Dimension 2",
               pch = 16,
               col = "blue")
        } 
        # 4. If the embedding dimension is 3 or more, create a 3D plot
        else if (embedding_dim >= 3) {
          scatterplot3d(
            x = ase$Xhat[, 1], 
            y = ase$Xhat[, 2], 
            z = ase$Xhat[, 3],
            main = well,
            xlab = "Dim 1",
            ylab = "Dim 2",
            zlab = "Dim 3",
            color = "blue",
            pch = 16,
            type = "p",
            grid = TRUE,
            box = TRUE
          )
        } 
        # Handle cases where the dimension is less than 2
        else {
          plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
          title(main = well)
          text(1, 1, paste("Embedding dim < 2\n(dim =", embedding_dim, ")"), col = "orange")
        }
      }
      
    }, error = function(e) {
      # If any other error occurs, plot a placeholder with the error message
      plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
      title(main = well)
      text(1, 1, "An error occurred", col = "red", cex = 0.8)
      message(paste("Error processing", well, ":", e$message))
    })
  }
  
  # --- Add the Main Title to the Entire Figure ---
  # Use mtext() to write the extracted path part in the outer margin (top)
  mtext(extract_path_part(file), 
        side = 3,       # 3 = top
        line = 1,       # Position in the margin
        outer = TRUE,   # Use the outer margin
        cex = 1.5,      # Character expansion (font size)
        font = 2)       # Font style (2 = bold)
  
  # Optional: Reset the plotting device layout to default
  # par(mfrow = c(1, 1), oma = c(0, 0, 0, 0))
  
  
}

```




